---
title: "GGIpack2: generic genome interrogation support"
shorttitle: "outlier dytection methods"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{GGIpack2: generic genome interrogation support}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

```{r setup,message=FALSE,echo=FALSE}
suppressWarnings({
library(GGIpack2)
library(duckdb)
library(arrow)
library(DBI)
library(dplyr)
library(EnsDb.Hsapiens.v75)
library(GenomicRanges)
library(DT)
})
```

This package addresses tasks of integrating and interrogating
diverse aspects of structural and functional aspects of
genomes, with initial focus on the genome of _Homo sapiens_.

We will show how to integrate information from GTEx (v7) and
Ensembl (v75) using resources in this package and in Bioconductor's
annotation facilities.

# Use with GTEx version 7 extracts

The following example shows how to obtain a slice of
data from GTEx eQTL data relating SNP genotypes
to gene expression in whole blood.

The
package has an example of data on eQTLs based on
RNA-seq analysis of whole blood samples.
The example data are 500 records from the GTEx
v7 summaries, filtered to keep only SNP loci
for which a test of association of rare
allele count with some nearby
gene had nominal p-value less than 0.05.

## Set up database connection and construct the interface

First define the path to a parquet file.  The
```{r lkparq, message=FALSE}
library(GGIpack2)
blex = system.file("parquet", "testbl.parquet", package="GGIpack2")
```

Second, set up a duckdb connection to facilitate use
of SQL-like querying of the data.

```{r docon}
con = DBI::dbConnect(duckdb::duckdb())
```

Third, use the `GTExresource` constructor to
create a systematic interface for querying.

```{r dogtr}
wb = GTExresource(con, tisstag="wholebl", pfile=blex)
wb
```

A helper method `getdb` simplifies direct access to the table.

## Explore the data and transform a slice to GRanges

We defined GTExresource to ensure that specific fields are 
available in the table associated with the object.

```{r lkn}
all(c("score", "seqnames", "start", "end") %in% colnames(getdb(wb)))
getdb(wb) |> group_by(seqnames) |> summarise(n=n())
```
That result implies that all the addresses are on chromosome 1.
We can assess multiplicities of  feature 'starts':

```{r lkmult}
getdb(wb) |> group_by(start) |> summarise(n=n()) |> arrange(desc(n))
```

To interpret these multiplicities, it is helpful to know which
genes (if any) are near the loci that are reported multiple times.
Conversion to GRanges is a good first step.

```{r doconv}
gr1 = toGRanges(getdb(wb) |> dplyr::filter(start %in% c(126113, 115746)))
genome(gr1) = "GRCh37"
gr1[, "molecular_trait_id"]
```


## Integrate with Ensembl annotation

To learn the gene names and biotypes of the molecular traits
identified above, use EnsDb.Hsapiens.v75 (for GRCh37).

```{r getens,message=FALSE}
library(EnsDb.Hsapiens.v75)
g37 = genes(EnsDb.Hsapiens.v75)
mults = g37[ gr1$molecular_trait_id ]
dd = as.data.frame(mcols(mults))

datatable(dd)
```


Finally, close the connection.

```{r dodisc}
DBI::dbDisconnect(con)
```

